---
output:
  md_document:
    variant: markdown_github
---

# Purpose

Purpose of this work folder.

Ideally store a minimum working example data set in data folder.

Add binary files in bin, and closed R functions in code. Human Readable settings files (e.g. csv) should be placed in settings/


```{r}

rm(list = ls()) # Clean your environment:
gc() # garbage collection - It can be useful to call gc after a large object has been removed, as this may prompt R to return memory to the operating system.
library(tidyverse)
list.files('code/', full.names = T, recursive = T) %>% .[grepl('.R', .)] %>% as.list() %>% walk(~source(.))
library(caTools)
library(caret)
library(glm2)

```

Looking at the 2016 Presidential Election
```{r}
#read in data
polls<-read.csv("/Users/ryanyoung/Desktop/Data Science/Machine Learning Project/Presidential Polls /pres_pollaverages_1968-2016.csv")
```


```{r}
#only look at national polls
polls <- polls[grepl("National", polls$state), ]
```

```{r}
#only look at Donald Trump and Hillary Clinton
polls <- polls[grepl("Donald Trump|Hillary Rodham Clinton", polls$candidate_name), ]

```

```{r}
# formating as date 
polls$modeldate <- as.Date(polls$modeldate, format = "%m/%d/%Y")

```

```{r}

#creating columns for Trump's poll rating and Clinton's poll rating 
polls_reformatted <- polls%>% mutate(Trump_poll = ifelse(candidate_name == "Donald Trump", pct_estimate, NA),
         Clinton_poll = ifelse(candidate_name == "Hillary Rodham Clinton", pct_estimate, NA))

#dropping unnecessary variables 
polls_reformatted <- polls_reformatted %>% select(state,Trump_poll, Clinton_poll, modeldate,pct_estimate)
library(dplyr)

#merging the date columns so that there is just one row per individual poll

polls_reformatted<- polls_reformatted %>%
  group_by(modeldate) %>%
  summarize(state = first(state),
            pct_estimate = first(pct_estimate),
            Trump_poll = max(Trump_poll, na.rm = TRUE),
            Clinton_poll = max(Clinton_poll, na.rm = TRUE))%>%ungroup() %>%
  select(-pct_estimate)



#create column for winning candidate
polls_reformatted <- polls_reformatted %>%
  mutate(winning_candidate = ifelse(Trump_poll > Clinton_poll, "Trump", 
                                    ifelse(Clinton_poll > Trump_poll, "Clinton", "Tie")))


#create binary variable called Clinton win 
polls_reformatted$ClintonWin <- ifelse(polls_reformatted$winning_candidate == 'Clinton', 1, 0)


```

 #Linear Regression Model
```{r}
#split data into training and testing set 

set.seed(123)
split <- sample.split(polls_reformatted$ClintonWin, SplitRatio = 0.7)
training_set <- subset(polls_reformatted, split == TRUE)
test_set <- subset(polls_reformatted, split == FALSE)
```

```{r}
# Train the model
model <- lm(ClintonWin ~ Trump_poll + Clinton_poll, data = training_set)
summary(model)
```

```{r}
# Predict outcomes
predictions <- predict(model, newdata = test_set)

```

```{r}
# Convert predictions to binary outcomes (threshold = 0.5)
predicted_outcome <- ifelse(predictions > 0.5, 1, 0)
print(predicted_outcome)
```

```{r}
# Calculate Accuracy
accuracy <- sum(predicted_outcome == test_set$ClintonWin) / nrow(test_set)
print(paste('Accuracy:', accuracy))
```

```{r}


# Create a confusion matrix
cm <- confusionMatrix(as.factor(predicted_outcome), as.factor(test_set$ClintonWin))

# The confusion matrix has entries for accuracy, precision (Positive Predictive Value), recall (Sensitivity)
# and F1 score (F1)
print(cm)

```

Logistic Regression Model 
```{r}


# Train the model
logistic_model <- glm(ClintonWin ~ Trump_poll + Clinton_poll, data = training_set, family = binomial())

# Print the summary
summary(logistic_model)

# Predict outcomes
predictions_logistic <- predict(logistic_model, newdata = test_set, type = "response")

# Convert to binary outcomes
predicted_outcome_logistic <- ifelse(predictions_logistic > 0.5, 1, 0)

# Calculate Accuracy
accuracy_logistic <- sum(predicted_outcome_logistic == test_set$ClintonWin) / nrow(test_set)
print(paste('Accuracy:', accuracy_logistic))

# Create a confusion matrix
cm_logistic <- confusionMatrix(as.factor(predicted_outcome_logistic), as.factor(test_set$ClintonWin))

# Print the confusion matrix
print(cm_logistic)

```

Looking at the 2024 Presidential Election 

```{r}
#read in data
polls_2024<-read.csv("/Users/ryanyoung/Desktop/Data Science/Machine Learning Project/Presidential Polls /president_polls_2024_cycle.csv")

```

```{r}
#cleaning the data
#only look at polling compare Trump and Biden, first two rows of each poll
polls_2024<- polls_2024%>%
    group_by(poll_id) %>%
    slice(1:2) %>%
    filter(all(c("Joe Biden", "Donald Trump") %in% candidate_name)) %>%
    ungroup()


#remove state level polls
polls_2024 <-polls_2024%>%
    filter(state == "")

#convert end_date into a date column
polls_2024$end_date <- as.Date(polls_2024$end_date, format = "%m/%d/%Y")

#convert pct in a numeric column 
polls_2024$pct <- as.numeric(polls_2024$pct)
```


```{r}
#creating columns for Trump's poll rating and Bidens poll rating 
polls_2024<- polls_2024%>% mutate(Trump_poll = ifelse(candidate_name == "Donald Trump", pct, NA),
         Biden_poll = ifelse(candidate_name == "Joe Biden", pct, NA))

#dropping unnecessary variables 
polls_2024 <- polls_2024%>% select(state, candidate_name, pollster,Trump_poll, Biden_poll, end_date,pct)
```

```{r}
#merging the date columns so that there is just one row per individual poll

polls_2024<- polls_2024 %>%
  group_by(end_date) %>%
  summarize(state = first(state),
            pct = first(pct),
            Trump_poll = max(Trump_poll, na.rm = TRUE),
            Biden_poll = max(Biden_poll, na.rm = TRUE))%>%ungroup() %>%
  select(-pct)
```

```{r}
#create column for winning candidate
polls_2024<- polls_2024 %>%
  mutate(winning_candidate = ifelse(Trump_poll > Biden_poll, "Trump", 
                                    ifelse(Biden_poll > Trump_poll, "Biden", "Tie")))


#create binary variable called Biden win 
polls_2024$BidenWin <- ifelse(polls_2024$winning_candidate == 'Biden', 1, 0)

```

Linear Regression Model 

```{r}
#split data into training and testing set 

set.seed(123)
split <- sample.split(polls_2024$BidenWin, SplitRatio = 0.7)
training_set_2024 <- subset(polls_2024, split == TRUE)
test_set_2024 <- subset(polls_2024, split == FALSE)
```

```{r}
# Train the model
model_2024 <- lm(BidenWin ~ Trump_poll + Biden_poll, data = training_set_2024)
summary(model_2024)
```

```{r}
# Predict outcomes
predictions_2024 <- predict(model_2024, newdata = test_set_2024)

```

```{r}
# Convert predictions to binary outcomes (threshold = 0.5)
predicted_outcome_2024 <- ifelse(predictions_2024 > 0.5, 1, 0)
print(predicted_outcome_2024)
```

```{r}
# Calculate Accuracy
accuracy_2024 <- sum(predicted_outcome_2024 == test_set_2024$BidenWin) / nrow(test_set_2024)
print(paste('Accuracy:', accuracy_2024))
```

```{r}
# Create a confusion matrix
cm_2024 <- confusionMatrix(as.factor(predicted_outcome_2024), as.factor(test_set_2024$BidenWin))

# The confusion matrix has entries for accuracy, precision (Positive Predictive Value), recall (Sensitivity)
# and F1 score (F1)
print(cm_2024)
```

#Logistic Regression Model

```{r}

# Train the model
logistic_model_2024 <- glm(BidenWin ~ Trump_poll + Biden_poll, data = training_set_2024, family = binomial())

# Print the summary
summary(logistic_model_2024)

# Predict outcomes
predictions_logistic_2024 <- predict(logistic_model_2024, newdata = test_set_2024, type = "response")

# Convert to binary outcomes
predicted_outcome_logistic_2024 <- ifelse(predictions_logistic_2024 > 0.5, 1, 0)

# Calculate Accuracy
accuracy_logistic_2024 <- sum(predicted_outcome_logistic_2024 == test_set_2024$BidenWin) / nrow(test_set_2024)
print(paste('Accuracy:', accuracy_logistic_2024))

# Create a confusion matrix
cm_logistic_2024 <- confusionMatrix(as.factor(predicted_outcome_logistic_2024), as.factor(test_set_2024$BidenWin))

# Print the confusion matrix
print(cm_logistic_2024)

```

Linear Model predicting the percentage of the vote scored 
```{r}
# Create a linear regression model to predict Biden's vote percentage
linear_model_biden <- lm(Biden_poll ~ Trump_poll, data = training_set_2024)

# Print summary
summary(linear_model_biden)

# Use the model to predict Biden's vote percentage in the test set
predicted_percentage_biden <- predict(linear_model_biden, newdata = test_set_2024)

# Calculate mean squared error or some other suitable metric
mse_biden <- mean((predicted_percentage_biden - test_set_2024$Biden_poll)^2)

# Repeat the process for Trump
linear_model_trump <- lm(Trump_poll ~ Biden_poll, data = training_set_2024)
summary(linear_model_trump)
predicted_percentage_trump <- predict(linear_model_trump, newdata = test_set_2024)
mse_trump <- mean((predicted_percentage_trump - test_set_2024$Trump_poll)^2)

```

